\pkg{partykit}: A Toolkit for Recursive Partytioning

### 2010-04-22 pkg/inst/doc/introduction.Rnw
tree example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@163 255f4249-2263-4417-b1ea-8498779e27a7

\author{Torsten Hothorn\\Ludwig-Maximilians-\\Universit\"at M\"unchen
   \And Achim Zeileis\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

@@

  Achim Zeileis\\
  Institut für Statistik \\
  Fakultät für Volkswirtschaft und Statistik \\
  Leopold-Franzens-Universität Innsbruck \\
  Universitätsstraße 15 \\
  6020 Innsbruck, Österreich
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://statmath.wu.ac.at/~zeileis/}
}

@@

@@

>>>>

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at M\"unchen
   \And Achim Zeileis\\Leopold-Franzens-Universität Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

@@

  Achim Zeileis\\
  Institut f\"ur Statistik \\
  Fakult\"at f\"ur Volkswirtschaft und Statistik \\
  Leopold-Franzens-Universit\"at Innsbruck \\
  Universit\"atsstra{\ss}e 15 \\
  6020 Innsbruck, \"Osterreich \\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://statmath.wu.ac.at/~zeileis/}
}

@@

\SweaveOpts{eps=FALSE, keep.source=TRUE, eval = TRUE}

@@

\section{My first Tree}

Package \pkg{partykit} does not offer unified infrastructure for growing trees. However,
once you know how to estimate splits from data, it is fairly straightforward to
implement trees. Consider a very simple tree algorithm. We assume that both response
and features are numeric. We search for the binary best split by means of $t$-test 
$p$-values, i.e., we cycle through all variables and potential split points and assess
the quality of the split by comparing the distributions of the response in the so-defined
two groups. We select the feature/split point combination with lowest two-sided $p$-value,
however only if this result is significant at level $\alpha = 0.05$.

This strategy can be implemented based on the data (response and features) and 
some case weights as follows:
<<mytree-1, echo = TRUE>>=
findsplit <- function(response, data, weights) {

    ### extract response values from data
    y <- data[[response]]

    logpmin <- 0
    xselect <- NULL

    ### cycle through all features
    for (i in which(names(data) != response)) {

        ### expand data
        x <- data[[i]]
        xt <- rep(x, weights)
        yt <- rep(y, weights)

        ### potential split points (not too many)
        qx <- unique(quantile(xt, 
                     prob = seq(from = 0.1, to = 0.9, by = 0.05)))

        ### assess all potential splits by their t-test
        ### log-p-value
        logp <- sapply(qx, function(q) {
            tt <- t.test(yt[xt <= q], yt[xt > q])
            pt(-abs(tt$statistic), tt$parameter, log = TRUE) + log(2)
        })

        ### if the best split in variable i significant AND
        ### better than what we already had store this information
        if (min(logp) < logpmin & min(logp) < log(0.05)) {
            logpmin <- min(logp)
            xselect <- i
            splitpoint <- qx[which.min(logp)]
        }
    }

    ### no significant split found, give up
    if (is.null(xselect)) return(NULL)

    ### return split as partysplit object
    return(partysplit(
        varid = as.integer(xselect),       ### which variable?
        breaks = as.numeric(splitpoint),   ### which split point?
        info = list(pvalue = exp(logpmin)  ### save p-value in addition
    )))
}
@

Now we have to set-up the recursion for growing a recursive 
\code{partynode} structure:
<<mytree-2, echo = TRUE>>=
growtree <- function(id = 1L, response, data, weights) {

    ### for less than 30 obs. stop here
    if (sum(weights) < 30) return(partynode(id = id))

    ### find best split
    sp <- findsplit(response, data, weights)
    ### no split found, stop here
    if (is.null(sp)) return(partynode(id = id))

    ### actually split the data
    kidids <- kidids_split(sp, data = data)

    ### set-up all daugther nodes
    kids <- vector(mode = "list", length = max(kidids))
    for (kidid in 1:max(kidids)) {
          ### select obs for current node
          w <- weights
          w[kidids != kidid] <- 0
          ### get next node id
          if (kidid > 1) {
              myid <- max(nodeids(kids[[kidid - 1]]))
          } else {
              myid <- id
          }
          ### start recursion on this daugther node
          kids[[kidid]] <- growtree(id = as.integer(myid + 1), 
              response, data, w)
    }

    ### return nodes
    return(partynode(id = as.integer(id), 
                     split = sp, kids = kids))
}
@

<<mytree-3, echo = TRUE>>=
mytree <- function(formula, data, weights = NULL) {

    ### name of the response variable
    response <- all.vars(formula)[1]
    ### data without missing values, response comes last
    data <- data[complete.cases(data), 
                 c(all.vars(formula)[-1], response)]
    if (is.null(weights)) weights <- rep(1, nrow(data))

    ### grow tree
    nodes <- growtree(id = 1L, response, data, weights)

    ### compute terminal node number for each obs.
    fitted <- fitted_node(nodes, data = data)
    ### return rich object
    ret <- party(nodes, 
        data = data,
        fitted = data.frame("(fitted)" = fitted,
                            "(response)" = data[[response]],
                            "(weights)" = weights,
                            check.names = FALSE),
        terms = terms(formula))
    as.constparty(ret)
}
@

<<mytree-4, echo = TRUE>>=
aqt <- mytree(Ozone ~ Solar.R + Wind + Temp, data = airquality)
aqt
@

\begin{figure}
\begin{center}
<<mytree-5, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
plot(aqt)
@
\caption{Tree}
\end{center}
\end{figure}

<<mytree-6, echo = TRUE>>=
predict(aqt, newdata = airquality[1:10,])
@

<<mytree-7, echo = TRUE>>=
aqt4 <- aqt[4]
aqt4
@

\begin{figure}
\begin{center}
<<mytree-8, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
plot(aqt4)
@
\caption{Subtree}
\end{center}
\end{figure}

<<mytree-9, echo = TRUE>>=
info_split(split_node(node_party(aqt4)))
@
